---
title: "Update histolgies file"
author: 'Antonia Chroni <chronia@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---


# Background
This is to update histologies file for the `pbta-tumor-evolution` project by: (1) adding `plot_group` column and (2) using `2023-11-13_histologies-base` in which approx. 150 RNA samples of poor quality have been removed. 

This module will be removed once v13 is released.

# Usage
To run the Rscript from the command line sequentially, use:

```
Rscript -e "rmarkdown::render('01_update_histologies.Rmd', clean = TRUE)"
```

# Set up
```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

## Directories and File Inputs/Outputs
```{r set-dir-and-file-names}
# Detect the ".git" folder -- this will be in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions no
# matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "update-histologies-temp")
input_dir <- file.path(analysis_dir, "input")

# Inputs
v12_file <- file.path(data_dir, "histologies.tsv")
updated_hist_file <- file.path(input_dir, "2023-11-13_histologies-base.tsv")
plot_mapping_file <- file.path(input_dir, "plot-mapping - plot-mapping.tsv")

# File path to results directory
results_dir <-
  file.path(analysis_dir, "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

# Read in data and process Data
```{r read-process-data, echo=TRUE}
v12_df <- readr::read_tsv(v12_file, guess_max = 100000, show_col_types = FALSE) %>% 
    select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, cancer_group, molecular_subtype, integrated_diagnosis, harmonized_diagnosis) 

updated_hist_df <- readr::read_tsv(updated_hist_file, guess_max = 100000, show_col_types = FALSE) %>% 
   select(- c(molecular_subtype, integrated_diagnosis, harmonized_diagnosis))

plot_mapping_df <- readr::read_tsv(plot_mapping_file, guess_max = 100000, show_col_types = FALSE) %>% 
  select(broad_histology, cancer_group, plot_group)

df <- plot_mapping_df %>%
  left_join(v12_df, by = join_by(cancer_group), relationship = "many-to-many") %>%
  dplyr::mutate(duplicates = if_else(plot_group == cancer_group, # | A == C | B == C
                              TRUE,
                              FALSE))%>% 
  filter(duplicates == TRUE) %>%
  select(!duplicates) %>%
  right_join(updated_hist_df, by = join_by(broad_histology, Kids_First_Participant_ID, Kids_First_Biospecimen_ID)) %>%
  mutate(
         # update 7316-3066
         broad_histology = case_when(sample_id == "7316-3066" ~ "Tumor of cranial and paraspinal nerves", 
                                     broad_histology == "Other" ~ "Other tumor",
                                     cancer_group == "Glial-neuronal tumor" ~ "Neuronal and mixed neuronal-glial tumor",
                                     cancer_group %in% c("Cavernoma", "Malignant peripheral nerve sheath tumor") ~ "Benign tumor",
                                     TRUE ~ broad_histology),
         cancer_group = case_when(sample_id == "7316-3066" ~ "Neurofibroma/Plexiform",
                                  grepl("xanthogranuloma", pathology_free_text_diagnosis) & broad_histology == "Histiocytic tumor" & pathology_diagnosis == "Other" ~ "Juvenile xanthogranuloma",
                                  broad_histology == "Choroid plexus tumor" ~ "Choroid plexus tumor",
                                  cancer_group == "Glial-neuronal tumor" ~ "Glial-neuronal tumor NOS",
                                  cancer_group == "Low-grade glioma" ~ "Low-grade glioma/astrocytoma",
                                  cancer_group %in% c("High-grade glioma", "Astrocytoma", "Astroblastoma", "Glioblastoma", "Diffuse hemispheric glioma",
                                                      "Infant-type hemispheric glioma") ~ "High-grade glioma/astrocytoma",
                                  broad_histology == "Ependymal tumor" ~ "Ependymoma",
                                  broad_histology == "Other tumor" ~ "Other tumor",
                                  broad_histology == "Diffuse astrocytic and oligodendroglial tumor" & (is.na(cancer_group) | cancer_group == "Oligodendroglioma") ~ "High-grade glioma/astrocytoma",
                                  cancer_group %in% c("Non-germinomatous germ cell tumor", "Diffuse leptomeningeal glioneuronal tumor", "Malignant peripheral nerve sheath tumor") ~ NA_character_,
                                  broad_histology == "Meningioma" ~ "Meningioma",
                                  cancer_group == "Perineuroma" ~ "Neurofibroma/Plexiform",
                                  is.na(cancer_group) & broad_histology == "Tumor of cranial and paraspinal nerves" ~ "Neurofibroma/Plexiform",
                                  TRUE ~ cancer_group)) %>% 
  write_tsv(file.path(results_dir, "histologies_2023_11_15.tsv")) 

```


```{r}
sessionInfo()
```
  