---
title: "Longitudinal analysis for the pbta-tumor-evolution project and PBTA Cohort"
author: 'Antonia Chroni <chronia@chop.edu> and Jo Lynne Rokita <rokita@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---


#### Tumor evolution project 

# Background
This is an exploratory analysis of longitudinal data in the PBTA cohort.
We are investigating the number of histologies (cancer_type column): (1) per assay (experimental_strategy) and (2) pairs of genomic and transcriptomic assays.
We are also looking at the number of patient samples available per histology with paired genomic and transcriptomic assays for (1) each time point (disease_stage) and (2) when Diagnosis is present and paired with another point in time.

# data used 
In this notebook, we are looking into the v12 histologies file (last updates on May 1st, 2023) and subset to the PBTA cohort.
We use tumor_descriptor and focus on disease_stages: Diagnosis, Progressive, Recurrence, and Deceased.

# note
The colors in the plots do not match the ones of the PBTA cohort yet. ---TO DO---


# Usage

This notebook is intended to be run via the command line from the top directory
of the repository as follows:

```
Rscript -e "rmarkdown::render('analyses/01_sample_distribution_analysis/01-tumor-descriptor-and-assay-count-plot.Rmd', clean = TRUE)"
```

# Set up

```{r load-library}

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(flextable)
  library(RColorBrewer) # to discard later
})


```

## Directories and File Inputs/Outputs

```{r set-dir-and-file-names}
# Detect the ".git" folder -- this will be in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions no
# matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)


# File path to input directory
input_dir <-
  file.path(root_dir, "input")


# Inputs
input_metadata <- file.path(input_dir, "histologies.tsv")
input_cols <- file.path(input_dir, "PBTA-germline-histology-groups-plot-mapping.tsv")


# File path to results directory
results_dir <-
  file.path(root_dir, "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}


# File path to plots directory
plots_dir <-
  file.path(root_dir, "plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}


# Outputs
output_PBTA_filter <- file.path(results_dir, "data_by_cohort_PBTA_filter.tsv")
output_PBTA_filter_count <- file.path(results_dir, "data_by_cohort_PBTA_filter_count.tsv")

# to add this later
# source publication theme
# source(file.path(root_dir, "figures", "theme.R"))
```


## Load data
```{r load-inputs-please-wait}
# Inputs
## Read in metadata
histologies <- readr::read_tsv(input_metadata, guess_max = 100000, show_col_types = FALSE)
# dim(data)

# Read table with colors per histology
cols <- readr::read_tsv(input_cols, guess_max = 100000, show_col_types = FALSE) # Load plotting colors

# Add colors palette to the histologies file
hist_cols <- histologies %>% left_join(cols, by=c('cancer_group','broad_histology'))  #to merge tables
# This is the tidyverse method. Try not to name variables the same as a function's name - that may cause coding errors downstream.


# to discard when add the col palette for PBTA cohort
mypalette <- brewer.pal(12,"Paired") # colorblindFriendly = TRUE
col5 <- colorRampPalette(mypalette)(5)
col9 <- colorRampPalette(mypalette)(9)

```

# Process Data

All tumor samples have matched normal samples. 
We need to filter normal samples out for further analysis. 
We are using "pathology_diagnosis" instead of "sample_type". This way we discard normal samples (as shown in "sample_type") in addition to undiagnosed tumor samples.
Also, we will discard Derived Cell Line, PDX from "composition".


```{r cohort PBTA, echo=TRUE}
data <- hist_cols  %>%
   filter(cohort == "PBTA",
          !is.na(pathology_diagnosis),
          !composition %in% c("Derived Cell Line", "PDX")) %>% 
         # !is.na(cancer_group) # Discard NAs in cancer type
   mutate(tumor_descriptor = ifelse(sample_id == "C4106355-Tumor", "Recurrence", # to be updated in v13 as "Recurrence"
                            ifelse(sample_id == "C4113612-Tumor", "Diagnosis", # to be updated in v13 as "Initial CNS Tumor"
                            ifelse(tumor_descriptor == "Primary Tumor", "Diagnosis",
                            ifelse(tumor_descriptor == "Initial CNS Tumor", "Diagnosis",
                            ifelse(tumor_descriptor == 'Progressive Disease Post-Mortem', 'Deceased',      
                                          tumor_descriptor))))))
                 
print (data %>% count(tumor_descriptor))


# Note
# The following samples (sample_id) are marked as Not Reported tumor_descriptor: 7316UP-3705, C4106355-Tumor, and C4113612-Tumor
# Jo Lynne found the following status for 2/3 in the data warehouse:
# C4106355-Tumor = Recurrence
# C4113612-Tumor = Initial CNS Tumor
# 7316UP-3705 = Not Reported
# manually annotate them here and these will be updated in the v13 - then discard that part of the code

# For now, you can leave this - we can fix the plot colors later - it might be that we have a longitudinal sample which comes from a benign tumor - we can possibly remove these later, but it would be good to get a source of truth for what our entire cohort looks like first before removing data,

# Save data after filter 
write_tsv(data, path = output_PBTA_filter)

# Coming back to this - there is also a possibility this is not reported or unknown - you will want to be sure these are captured.
# If this is the case for any patient with longitudinal data, let's go over these and we can fill the info in from the clinical team.

# Can you comment around what you are doing here? I think you might want to use fewer columns and not use sample id here - sample_id is tied to an event, so if a patient has multiple events, then you would lose that information here. But maybe you are not trying to count that yet?

count1 <- data %>% 
          group_by(cohort_participant_id, Kids_First_Participant_ID,
                 experimental_strategy, sample_type, tumor_descriptor, primary_site,
                 age_at_diagnosis_days, age_last_update_days, age_at_chemo_start, age_at_radiation_start,
                 pathology_diagnosis, cancer_group, broad_histology) %>% 
          dplyr::count(cancer_group) 

write_tsv(count1, output_PBTA_filter_count)

```

# Plots
## All assays

```{r plot1}
count1$experimental_strategy <- factor(x = count1$experimental_strategy, levels = c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing", "Methylation"))

p <- ggplot(count1, aes(x=cancer_group, fill=experimental_strategy))+
        geom_bar()+
        scale_fill_manual(values=col5) +
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs experimental_strategy") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave("Barplot_cancer_group_experimental_strategy_PBTA.pdf", plot=p, path=plots_dir, device="pdf", height=7, width=10)

```

## Paired genomic, transcriptomic assays

```{r plot2}
count <- count1 %>% filter(experimental_strategy %in% c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing")) # Not included:Methylation
count$experimental_strategy <- factor(x = count$experimental_strategy, levels = c("WGS", "WXS", "RNA-Seq", "Targeted Sequencing"))

p <- ggplot(count, aes(x=cancer_group, fill=experimental_strategy))+
        geom_bar()+
        scale_fill_manual(values=col5) + #scale_fill_manual(values=c(data$plot_group_hex)) 
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs experimental_strategy") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave("Barplot_cancer_group_experimental_strategy_PBTA_matched_DNA-RNA.pdf", plot=p, path=plots_dir, device="pdf", height=7, width=10)

```

## By cancer_group - all disease stages


```{r plot3}
count$tumor_descriptor <- factor(x = count$tumor_descriptor, levels = c("Diagnosis", "Progressive", "Recurrence", "Deceased", "Second Malignancy", "Metastatic", "Not Reported", "Residual", "Unavailable"))


p <- ggplot(count, aes(x=cancer_group, fill=tumor_descriptor))+
        geom_bar()+
        scale_fill_manual(values=col9) +
        theme_classic()+
        ylab("Count cancer_group") +
        ggtitle("cancer_group vs tumor_descriptor") +
        guides(color=guide_legend(override.aes=list(size=3))) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
        print(p)
ggsave("Barplot_cancer_group_tumor_descriptor_PBTA_matched_DNA-RNA.pdf", plot=p, path=plots_dir, device="pdf", height=7, width=10)

```

# Summary

In the following section, we will provide summary table across different disease stages, histologies, and patients samples.

### Number of tumor samples per each assay

First, we’ll examine how many of each type of assay we have (tumors only). This information is stored in the experimental_strategy column.

```{r}
count1 %>%
  group_by(experimental_strategy) %>%
  tally() %>%
  arrange(desc(n)) %>% 
  regulartable() %>%
  fontsize(size = 12, part = "all")

```



### Paired genomic, transcriptomic assays

#### Multiple time points: Diagnosis, Progressive, Recurrence, Deceased
We're going to use the `tumor_descriptor` column here with matched genomic and transcriptomic assays from before.


#### By Kids_First_Participant_ID
### Disease stage - descriptors pairs

How many tumor samples per disease_stage?

```{r}
disease_stage <- count %>%
  group_by(Kids_First_Participant_ID) %>%
  summarize(disease_stage = paste(sort(unique(tumor_descriptor)),
                                  collapse = ", "),
                                collapse = ", ") %>%
  group_by(disease_stage) %>%
  tally() %>%
  arrange(desc(n))

disease_stage  %>% 
  regulartable() %>%
  fontsize(size = 12, part = "all")

```


What about when Diagnosis is present and paired with another point in time?

```{r}
Diagnosis_paired <- disease_stage  %>%
  filter(disease_stage != "Diagnosis")  %>% 
  filter(grepl("Diagnosis", disease_stage)) %>%
  regulartable() %>%
  fontsize(size = 12, part = "all")
Diagnosis_paired
```


