---
title: "Create VAF corplots of tumors across multiple timepoints for thr autopsy samples in the PBTA Cohort"
author: 'Antonia Chroni <chronia@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

#### Tumor evolution project 

### Data used 
In this notebook, we are using the `maf_autopsy.tsv` file generated from the `01-preprocess-data.Rmd` script.
We will also use the oncoprint goi list from OpenPedCan `oncoprint-goi-lists-OpenPedCan-gencode-v39.csv`.

# Set up

```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggrepel)
})
```

## Directories and File Inputs/Outputs

```{r set-dir-and-file-names}
# Detect the ".git" folder. This will be in the project root directory.
# Use this as the root directory to ensure proper sourcing of functions
# no matter where this is called from.
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
scratch_dir <- file.path(root_dir, "scratch")
analysis_dir <- file.path(root_dir, "analyses", "tmb-vaf-longitudinal") 
input_dir <- file.path(analysis_dir, "input")

# Input files
maf_autopsy_file <- file.path(scratch_dir, "maf_autopsy.tsv")
oncoprint_goi_file <- file.path(input_dir, "oncoprint-goi-lists-OpenPedCan-gencode-v39.csv")
palette_file <- file.path(root_dir, "figures", "palettes", "timepoint_color_palette.tsv")

# File path to plots directory
plots_dir <-
  file.path(analysis_dir, "plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# File path to results directory
results_dir <-
  file.path(analysis_dir, "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

source(paste0(analysis_dir, "/util/function-create-corplot.R"))
source(paste0(root_dir, "/figures/scripts/theme.R"))
```

## Read in data and process
```{r read-input-files}
# Read maf_autopsy file generated from step `01-process-data.Rmd`
maf_autopsy_raw <- readr::read_tsv(maf_autopsy_file, guess_max = 100000, show_col_types = FALSE) %>% 
  group_by(Kids_First_Participant_ID) %>% 
  mutate(cg_distinct = n_distinct(cancer_group) > 1) # to identify samples with different diagnosis across timepoints

# So far we have one PT_1H2REHT2 patient case with a secondary diagnosis in one of the progressive samples
# To reflect situations with multiple diagnosis, we will create `cg_sum column`
maf_autopsy <- maf_autopsy_raw %>% 
  distinct(cancer_group, .keep_all = TRUE) %>% 
  summarise(cg_sum = str_c(cancer_group, collapse = ",")) %>%
  left_join(maf_autopsy_raw, by = c("Kids_First_Participant_ID"))

# Read oncoprint_goi_file
# We will use the following list of genes and then we will create a label to select the ones for the corplots.
oncoprint_goi <- read.csv(oncoprint_goi_file, stringsAsFactor = FALSE) %>%
  select(LGAT, Embryonal.tumor, HGAT, Other) 

# let's create a list with the genes to be used for the corplots
oncoprint_goi <- data.frame(goi_list = unlist(oncoprint_goi)) 

# remove empty rows
oncoprint_goi <- oncoprint_goi[!apply(oncoprint_goi == "", 1, all), ] 
oncoprint_goi <- data.frame(oncoprint_goi)

# Read palette_df file
palette_df <- readr::read_tsv(palette_file, guess_max = 100000, show_col_types = FALSE)
```

## Corplots for each Patient case and per biospecimen_id and timepoint

```{r create-corplot-please-wait}
# Save results by cg_sum
cg_names <- unique(as.character(maf_autopsy$cg_sum))
cg_names <- sort(cg_names, decreasing = FALSE)
print(cg_names)

# Loop through cg_sum
for (x in 1:length(cg_names)){
  print(x)

  cg_maf_sub <- maf_autopsy %>%
    filter(cg_sum == cg_names[x])
 
  # We will define samples based on the "Kids_First_Participant_ID" column 
  samples <- unique(as.character(cg_maf_sub$Kids_First_Participant_ID))
  print(samples)
    
  # define cg_sum specific plots dir
  cg_plots_dir <- file.path(plots_dir, cg_names[x])
  if (!dir.exists(cg_plots_dir)) {
    dir.create(cg_plots_dir)
  }
    
  # define cg_sum specific results dir
  cg_results_dir <- file.path(results_dir, cg_names[x])
  if (!dir.exists(cg_results_dir)) {
    dir.create(cg_results_dir)
  }
  
  # Run loop for each sample and pair
  for (i in seq_along(samples)) {
    print(i)
    maf_sub <- cg_maf_sub %>%
      filter(Kids_First_Participant_ID == samples[i])
    timepoints_other_plots <- unique(maf_sub$timepoints_other)
    timepoints_other_plots <- timepoints_other_plots[!timepoints_other_plots == "4"]
    print(timepoints_other_plots)
    timepoints_deceased_plots <- unique(maf_sub$timepoints_deceased)
    timepoints_deceased_plots <- timepoints_deceased_plots[!timepoints_deceased_plots %in% c("1", "2", "3")]
    print(timepoints_deceased_plots)
  
    # Run corplot function
    for (t in seq_along(timepoints_deceased_plots)){
      for (k in seq_along(timepoints_other_plots)){
        fname <- paste0(cg_plots_dir, "/", samples[i], "-", timepoints_other_plots[k], "-vs-", timepoints_deceased_plots[t], "-vaf-corplot.pdf")
        print(fname)
        p <- create_corplot(maf = maf_sub,
                            timepoints_other_plot = timepoints_other_plots[k],
                            timepoints_deceased_plot = timepoints_deceased_plots[t],
                            sid = samples[i])
        pdf(file = fname, width = 10, height = 8)
        print(p)
        dev.off()
      }
    }
  }  
}
```

```{r echo=TRUE}
sessionInfo()
```

