---
title: "TMB across all longitudinal samples in the PBTA Cohort"
author: 'Antonia Chroni <chronia@chop.edu> for D3B'
date: "2023"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
---

# Set up
```{r load-library}
suppressPackageStartupMessages({
  library(tidyverse)
  library(rstatix)
  library(ggpubr)
})
```

## Directories and File Inputs/Outputs
```{r set-dir-and-file-names}
# Detect the ".git" folder -- this will be in the project root directory
# Use this as the root directory to ensure proper sourcing of functions 
# no matter where this is called from
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "tmb-vaf-longitudinal")
input_dir <- file.path(analysis_dir, "input")
files_dir <- file.path(root_dir, "analyses", "sample-distribution-analysis", "results")

# Input files
pbta_file <- file.path(files_dir, "pbta.tsv") # file from add-sample-distribution module
tmb_file <- file.path(input_dir, "snv-mutation-tmb-coding.tsv")
palette_file <- file.path(root_dir, "figures", "palettes", "tumor_descriptor_color_palette.tsv")

# File path to plot directory
plots_dir <-
  file.path(analysis_dir, "plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# File path to results directory
results_dir <-
  file.path(analysis_dir, "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

source(paste0(root_dir, "/figures/scripts/theme.R"))
```

## Read in data and process
```{r load-process-inputs}
# Make this reproducible
set.seed(2023)

# Define timepoints
timepoints = c("Diagnosis", "Progressive", "Recurrence", "Deceased", "Second Malignancy", "Unavailable")

# Read in tmb file and remove hypermutant samples
tmb <- readr::read_tsv(tmb_file, guess_max = 100000, show_col_types = FALSE) %>%
  filter(!tmb >= 10) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID = Tumor_Sample_Barcode) # change name of the biospecimen to match the one from the histologies file

# Read in pbta file and create df
df <- readr::read_tsv(pbta_file, guess_max = 100000, show_col_types = FALSE) %>%
  filter(!(experimental_strategy == "RNA-Seq"),
         !(tumor_descriptor == "Unavailable")) %>% # these are 2 samples
  left_join(tmb, by = c("Kids_First_Biospecimen_ID", "experimental_strategy")) %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, tumor_descriptor,
         broad_histology, match_id_DNA, experimental_strategy, 
         cg_multiple, cg_id, cgGFAC, cancer_group, short_histology, tmb, mutation_count, DNA) %>%
  dplyr::mutate(log10_tmb = log10(tmb)) %>% 
  distinct(match_id_DNA, .keep_all = TRUE) %>% # keep only one bs_sample per genomic assay (WGS/WXS)
  filter(!is.na(tmb))

# How many samples per Timepoint?
print(df %>% dplyr::count(tumor_descriptor) %>% arrange(desc(n)))

# How many samples per cancer group?
print(df %>% dplyr::count(cg_id) %>% arrange(desc(n)))

# How many samples per Timepoint and cancer type?
# rename(count(df, tumor_descriptor, cg_id), Freq = n)

# How many cases with unpaired longitudinal samples?
no_samples <- length(unique(df$Kids_First_Participant_ID))

# Summarize timepoints per cancer groups
# We will summarize cancer groups with single timepoints as "Other" 
timepoint_sum <- df %>%
  distinct(match_id_DNA, .keep_all = TRUE) %>%
  group_by(cg_id) %>%
  arrange(tumor_descriptor) %>%
  summarise(tumor_descriptor_sum = str_c(tumor_descriptor, collapse = "-")) 

# Create df to use for plots
df <- df %>% 
  left_join(timepoint_sum, by = c("cg_id")) %>%
  filter(!tumor_descriptor_sum %in% timepoints,
         !tumor_descriptor_sum %in% c("Diagnosis-Diagnosis", 
                                      "Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis",
                                      "Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis",
                                      "Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis-Diagnosis"
                                      )) # remove cancer groups with single timepoints


# Let's count #samples per cancer groups and timepoints 
timepoint_cgGFAC_n_df <- df %>% 
  dplyr::count(cgGFAC, tumor_descriptor) %>% 
  dplyr::mutate(timepoint_cgGFAC_n = glue::glue("{tumor_descriptor} (N={n})")) %>% 
  dplyr::rename(timepoint_cgGFAC_number = n) 


# Create df to use for plots
df_plot <- df %>% 
  left_join(timepoint_cgGFAC_n_df, by = c("tumor_descriptor", "cgGFAC")) %>% 
  filter(!timepoint_cgGFAC_number <= 2) %>% # remove if total #samples per timepoint is less than 2
  mutate(cg_id_sum = replace_na(cg_id, "Other"),
         tumor_descriptor = factor(tumor_descriptor),
         tumor_descriptor = fct_relevel(tumor_descriptor, timepoints),
        
         # Edit name of `cg_id` to fit in the plot title for the panel
         cg_id_sum = str_wrap(cg_id_sum, 12)) 
  
``` 


## How many patient samples in the unpaired longitudinal data?
There are `r no_samples` samples in total.

# All unpaired samples
We are interested in how TMB changes over the time in all PBTA samples (unpaired samples).

```{r define-parameters-for-plots}
# Read color palette
palette_df <- readr::read_tsv(palette_file, guess_max = 100000, show_col_types = FALSE) %>% 
  mutate(tumor_descriptor = color_names)

# Define and order palette
palette <- palette_df$hex_codes
names(palette) <- palette_df$tumor_descriptor

# Define label for plots
Timepoint <- df_plot$tumor_descriptor

# Define ylim
ylim <- max(df_plot$tmb)
log10_ylim_min <- min(df_plot$log10_tmb)
log10_ylim_max <- max(df_plot$log10_tmb)
```

```{r plot-total, fig.width = 5, fig.height = 5, fig.fullwidth = TRUE}
# Make this reproducible
set.seed(2023)

print(ggplot(df_plot, aes(tumor_descriptor, tmb, color = Timepoint)) + 
        geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) + 
        # light guiding line representing 0 exposure
        geom_hline(yintercept = 0, size = 0.15) + 
        scale_color_manual(values = palette, 
                           breaks = sort(names(palette))) +
        stat_summary(fun.data = mean_se, color = "black", size = 0.3) + 
        labs(x = "Tumor descriptor",
             y = "TMB") +
        theme_Publication() +
        scale_y_continuous(limits = c(0, ylim)) +
        theme(axis.text.x = element_text(angle = 90)))

# Save the plot
ggsave(filename = "TMB-unpaired-longitudinal.pdf",
       path = plots_dir, 
       width = 5, 
       height = 5, 
       device = "pdf", 
       useDingbats = FALSE)
```


# All unpaired samples across cancer types

```{r plot-cg-id-sum, fig.width = 15, fig.height = 15, fig.fullwidth = TRUE}
# Make this reproducible
set.seed(2023)

print(ggplot(df_plot, aes(tumor_descriptor, log10_tmb, color = Timepoint)) + 
        geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) + 
        # light guiding line representing 0 exposure
        geom_hline(yintercept = 0, size = 0.15) + 
        scale_color_manual(values = palette, 
                           breaks = sort(names(palette))) +
        stat_summary(fun.data = mean_se, color = "black", size = 0.3) + 
        ggplot2::facet_wrap(~cg_id_sum, nrow = 8) +
        labs(x = "Tumor descriptor",
             y = "log10TMB") +
        theme_Publication() +
        scale_y_continuous(limits = c(log10_ylim_min, log10_ylim_max)) + #oob = scales::oob_keep
        theme(axis.text.x = element_text(angle = 90)))

# Save the plot
ggsave(filename = "TMB-unpaired-longitudinal-cg-id-sum.pdf", 
       path = plots_dir, 
       width = 15, 
       height = 15, 
       device = "pdf", 
       useDingbats = FALSE)
```

# Summary statistics for Timepoints and cancer types
Considering the inequalities in the sampling effort for each timepoint, let's explore whether the sampling size affects the observed patterns, e.g., indication of Progressive and Recurrence having higer TMB compared to Diagnosis for High-grade glioma patients.

```{r summary-statistics, fig.width = 8, fig.height = 6, fig.fullwidth = TRUE}
# Inspect some random rows of the data by groups
set.seed(2023)

df_plot %>% 
  select(cgGFAC, timepoint_cgGFAC_n, log10_tmb) %>% 
  sample_n_by(cgGFAC, timepoint_cgGFAC_n, log10_tmb, size = 1) 

# Define cancer_groups
cancer_groups <- unique(as.character(df_plot$cgGFAC))
cancer_groups <- sort(cancer_groups, decreasing = FALSE)
print(cancer_groups)


for (i in seq_along(cancer_groups)) {
  print(i)
  
  df_sub <- df_plot %>% 
    filter(cgGFAC == cancer_groups[i]) %>% 
    arrange(tumor_descriptor, timepoint_cgGFAC_n) 
 
  # To add cancer group in the stat.test table
  df_sub_list <- df_sub %>% 
    select(cgGFAC, tumor_descriptor, timepoint_cgGFAC_n) %>% 
    unique() %>% 
    arrange(tumor_descriptor, timepoint_cgGFAC_n) 
  
  # Compute some summary statistics (mean and sd) by groups:
  stat.summary <- df_sub %>%
    group_by(cgGFAC, timepoint_cgGFAC_n) %>%
    get_summary_stats(log10_tmb, type = "median_iqr") %>%
    as.data.frame() 

  stat.test <- df_sub %>% 
    rstatix::wilcox_test(log10_tmb ~ timepoint_cgGFAC_n) %>%
    add_significance() %>%
    as.data.frame() %>% 
    mutate(timepoint_cgGFAC_n = group1) %>% 
    left_join(df_sub_list, by = c("timepoint_cgGFAC_n")) %>% 
    arrange(cgGFAC, tumor_descriptor, timepoint_cgGFAC_n) %>% 
    select(-timepoint_cgGFAC_n)

  # Add add_xy_position
  stat.test <- stat.test %>% 
    add_y_position(step.increase = 0.5) #step.increase = 0.2
  
  # Save results for all 
  if(i == 1){
    stat.summary.all <- stat.summary
    stat.test.all <- stat.test
    } else {
      stat.summary.all <- rbind(stat.summary.all, stat.summary)
      stat.test.all <- rbind(stat.test.all, stat.test)
    }
  
  df_sub <- df_sub %>% 
    dplyr::mutate(timepoint_cgGFAC_n = forcats::fct_reorder(timepoint_cgGFAC_n, timepoint_cgGFAC_number, .desc=T))
  
    # Create bxp Wilcox-test
    p <- print(df_sub %>%  
                 ggplot2::ggplot() +
                 aes(x = timepoint_cgGFAC_n, y = log10_tmb, color = tumor_descriptor) + 
                 geom_jitter(width = 0.15, size = 0.7, alpha = 0.6) +
                 ggplot2::geom_boxplot(color = "black",
                              size = 0.25,
                              alpha = 0,
                              coef = 0) + # remove whiskers
          theme_Publication() + 
          scale_color_manual(values = palette) +
          #rotate() +
          theme(axis.text.x = element_text(angle = 90)) +
          stat_pvalue_manual(stat.test,
                             #tip.length = 0.01,
                             hide.ns = TRUE, 
                             label = "Wilcox-test, p = {p.adj}", 
                             #label = "p.adj.signif", 
                             y.position = "y.position") +
          scale_y_continuous(limits = c(log10_ylim_min, max(stat.test$y.position))) +
          labs(title = cancer_groups[i],
               x = "Tumor descriptor",
               y = "log10TMB",
               color = "Timepoint"))
    
    
  # Save the plot
  ggsave(filename = paste0(cancer_groups[i], "-TMB-Bxp-stat-test-Wilcox-test.pdf"), 
         path = plots_dir, 
         width = 8, 
         height = 6, 
         device = "pdf", 
        useDingbats = FALSE)
}


# Save summary tables for statistics
stat.summary.all %>% write_tsv(file.path(results_dir, "stat_summary.tsv"))
stat.test.all %>% write_tsv(file.path(results_dir, "stat_test.tsv"))
```

```{r echo=TRUE}
sessionInfo()
```
